# Results

## Overall Draft Analysis

### Points, Assists, Rebounds, Win Shares vs Pick Number (1-60)

```{r}
library(tidyr)

player_achievements <- read_csv("player_achievements.csv")
nbaplayersdraft <- read_csv("nbaplayersdraft.csv")

# Calculate the average for each metric by overall_pick
average_stats_by_pick <- nbaplayersdraft %>%
  filter(overall_pick <= 54) %>%
  group_by(overall_pick) %>%
  summarise(
    average_points_per_game = mean(points_per_game, na.rm = TRUE),
    average_total_rebounds = mean(average_total_rebounds, na.rm = TRUE),
    average_assists = mean(average_assists, na.rm = TRUE),
  )

# Convert to long format for faceting
long_stats_by_pick <- average_stats_by_pick %>%
  pivot_longer(
    cols = starts_with("average_"),
    names_to = "Metric",
    values_to = "Average"
  )

# Create the faceted plot
ggplot(long_stats_by_pick, aes(x = overall_pick, y = Average)) +
  geom_line() + 
  facet_wrap(~Metric, scales = "free_y", nrow = 4) +
  theme_minimal() +
  labs(title = "Average Player Stats by Draft Pick",
       x = "Overall Draft Pick",
       y = "Average") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.position = "none")


```

For each metric, they seem to level off around the 10-15th pick. So it would be to a team's benefit to get draft pick somewhere in the top 10. While there seems to be an occasional star player picked in between 15 and 40, generally a team shouldn't worry about the order as much.


### Advanced NBA Stats for Top 5 Picks Over time

```{r}

# Filter the data to include only the top 5 picks
top_picks <- nbaplayersdraft %>%
  filter(overall_pick >= 1 & overall_pick <= 5)

# Reshape the data to long format for faceting by metric
top_picks_long <- top_picks %>%
  select(year, overall_pick, win_shares, value_over_replacement) %>%
  pivot_longer(
    cols = c(win_shares, value_over_replacement),
    names_to = "Metric",
    values_to = "Value"
  )

# Create the faceted scatter plot for Win Shares and Value Over Replacement over time for the top 5 picks
ggplot(top_picks_long, aes(x = year, y = Value, color = factor(overall_pick))) +
  geom_point() +
  scale_color_brewer(palette = "Set1") +  # Use a qualitative color palette
  facet_wrap(~Metric, scales = "free_y") +  # Create facets for each metric
  theme_minimal() +
  labs(title = "Win Shares and Value Over Replacement for Top 5 Overall Picks Over Time",
       x = "Year",
       y = "Metric Value",
       color = "Overall Pick") +
  theme(legend.position = "right")

```

The first overall pick doesn't consistently end up being the best out of their draft class.


## Total Player Achievements and their Draft Team


```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)

merged_achievement_data <- left_join(player_achievements, nbaplayersdraft, by = c("Player Name" = "player"))

merged_achievement_data <- merged_achievement_data %>%
  filter(!is.na(team))

team_achievements <- merged_achievement_data %>%
  group_by(team) %>%
  summarise(
    TotalAllStarSelections = sum(Total_All_Star_Selections, na.rm = TRUE),
    TotalAllNBASelections = sum(Total_All_NBA_Selections, na.rm = TRUE),
    TotalChampionships = sum(Total_Championships, na.rm = TRUE)
  ) %>%
  arrange(desc(TotalAllStarSelections))

# Create separate Cleveland dot plots for each achievement
plot1 <- ggplot(team_achievements, aes(x = TotalAllStarSelections, y = reorder(team, TotalAllStarSelections))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Total All-Star Selections per Team",
       x = "Total All-Star Selections",
       y = "Team") +
  theme(axis.text.y = element_text(hjust = 1))

plot2 <- ggplot(team_achievements, aes(x = TotalAllNBASelections, y = reorder(team, TotalAllNBASelections))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Total All-NBA Selections per Team",
       x = "Total All-NBA Selections",
       y = "Team") +
  theme(axis.text.y = element_text(hjust = 1))

plot3 <- ggplot(team_achievements, aes(x = TotalChampionships, y = reorder(team, TotalChampionships))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Total Championships per Team",
       x = "Total Championships",
       y = "Team") +
  theme(axis.text.y = element_text(hjust = 1))

plot1
plot2
plot3

```



## College Draft Targets

### Points, Rebounds, Assists for NBA Players from each College


```{r}

# Calculate the totals for each metric by college
total_stats_by_college <- nbaplayersdraft %>%
  group_by(college) %>%
  summarise(
    total_points_per_game = sum(points_per_game, na.rm = TRUE),
    total_total_rebounds = sum(average_total_rebounds, na.rm = TRUE),
    total_assists = sum(average_assists, na.rm = TRUE)
  ) %>%
  filter(!is.na(college) & college != "") %>%
  ungroup()

# Function to create a Cleveland dot plot for a given metric
create_cleveland_plot <- function(data, metric, title) {
  top_colleges <- data %>%
    top_n(50, !!sym(metric))

  ggplot(top_colleges, aes(x = !!sym(metric), y = reorder(college, !!sym(metric)))) +
    geom_point() +
    theme_minimal() +
    labs(title = title, x = "Total", y = "College") +
    theme(axis.text.y = element_text(hjust = 1))
}

plot1 <- create_cleveland_plot(total_stats_by_college, "total_points_per_game", "Top 50 Colleges by Total Points Per Game")
plot2 <- create_cleveland_plot(total_stats_by_college, "total_total_rebounds", "Top 50 Colleges by Total Total Rebounds")
plot3 <- create_cleveland_plot(total_stats_by_college, "total_assists", "Top 50 Colleges by Total Assists")

plot1
plot2
plot3


```

### Top College Targets for Draft Teams

These colleges seem to be producing the best NBA players: Kentucky, Duke, UCLA, Arizona, Kansas. Are there team targeting players from these colleges on draft night?

```{r}

# Filter the dataset for the specified colleges
filtered_data <- nbaplayersdraft %>%
  filter(college %in% c("Kentucky", "Duke", "UCLA", "Arizona", "Kansas"))

# Count the number of players drafted by each team from these colleges
team_draft_counts <- filtered_data %>%
  count(team, college) %>%
  arrange(desc(n))

# Create a bar plot
ggplot(team_draft_counts, aes(x = reorder(team, n), y = n, fill = college)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "NBA Team Draft Counts from Top 5 Basketball Colleges",
       x = "Team",
       y = "Number of Players Drafted") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.position = "bottom")

```



## Drafting Based on Physical Attributes

```{r}
#install.packages("GGally")
library(GGally)
library(dplyr)


df_1 <- df[rowSums(is.na(df)) <= 10, ]

df_processed <- df_1 %>%
  mutate_if(is.numeric, ~ifelse(is.na(.), 0, .))

```

```{r}
df_para <- df_processed[, c(3, 4, 6:11, 22:34)]
df_related <- df_para[, c(1,2, 4,5, 8,12, 16, 17)]
#ggparcoord(df_para, alphaLines = .3,
#          scale = "uniminmax")
ggpairs(df_related,
        lower = list(continuous = wrap("points", alpha = 0.1, size=0.2)))
cor_matrix <- cor(df_para)
print(cor_matrix)
```
To analyze the factor of weight and height in drafting. We first analyzed the correlation matrix of the height and weight to the other performance statistics. It is reasonable that height and weight are highly correlated and statistics about rebound, assist, and X_3 points have high absolute value of correlation score with height and weight. Then, we analyze the pair plot of these variables. The distribution of performance variables are all positive skewed, while the distribution of height and weight are close to the normal distribution. The scatter plot of height and weight with performance statistics are almost the same. We have the same conclusion for them. The different thing is height plot tend to have grid shape, which may caused by players tend to round up their height to integer. Statistics about rebound are positive related to the height weight, Players with high weight and height tend to have larger variance in the rebound statistics, whereas player with low weight and height has lower variance. We can draw the following conclusion that if the team need to strength their rebound aspects, they should absolutely draw players with high height and weight, but these players may not get enough rebound in the match.The similar things happen to the assist statistics. Assist statistics are negatively correlated to the height and weight. Players with high weight and height tend to have lower variance in the assist statistics, whereas player with low weight and height has higher variance. If a team need to strength their team corporation(assist), they should absolutely draw players with low height and weight, but these players may not assist enough in the match. We also discover that the variance difference is more obvious in the average assist, but less in assist percentage(Percentage of teammate field goals the player assisted). ????any conclusion?

```{r}
df_origin <- df_processed %>%
  mutate(country = ifelse(country == "USA", "USA","non-American"))%>%
  group_by(country)%>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))%>%
  select(-na_count, -year, -id, -minutes_played, -points, -total_rebounds, -box_plus_minus, - win_shares_per_48_minutes, -win_shares,-off_rebound, -def_rebound,-usg_pct, -total_pointdiff)%>%
  lapply(function(x) if(is.numeric(x)) x / x[1] else x) %>%
  as.data.frame()%>%
  pivot_longer(cols = draft_number:value_over_replacement, names_to = "statistics" )
ggplot(df_origin, aes(statistics, value, fill = country))+
  geom_bar(stat="identity", position = position_dodge())
```
No difference observed

```{r}
#install.packages("hexbin")
df_1round <- df_processed |>
  select(player_height, player_weight, draft_round)
ggplot(df_1round, aes(x=player_height, y=player_weight))+
  geom_hex(bins = 10)+
  scale_fill_viridis_c()
```
The player with 200-210? eight and 90-110? weight conclusion?
```{r}
df_cleve <- df_processed %>%
  select(college, usg_pct:ast_pct, field_goal_percentage:free_throw_percentage)%>%
  group_by(college)%>%
  filter(n()>12)%>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))%>%
  arrange(free_throw_percentage) %>%  
  mutate(college = factor(college, levels = college)) %>%
  pivot_longer(cols=usg_pct:free_throw_percentage)
ggplot(df_cleve, aes(x = value, y = college, fill=name)) + 
  geom_point(stat = "identity", shape = 21, color = "black", size = 1.5) +
  labs(y = "College", x = "Category")+
  scale_color_manual(values = c("blue",  "red",  "green", "yellow", "purple", "#008080"))
```
To analyze the influence of college in draft, we analyze the players' performance in assist, field goal, free throw, shooting efficiency, and, team plays. The college will influence the style of the player in the shooting and team plays. This analysis is to find out the styles of players from certain college. To avoid the randomness of the players, we only analyze the colleges with more than 12 players. We noticed that there is significant difference in ts, usg, and field goals. Players from Gonzaga, Villanova, and Oregan tend to have higher free throw percentage, while players from Louisville, Syracuse, and Florida, tend to lower free throw percentage. If NBA team need a player to earn points through free throw, they should pick players from Louisville, Syracuse, and Florida. The players in performance in X_3 point is around the same, except players from Louisville, Florida, Maryland, and LSU. The only two college with outstanding X_3 point are Villanova, and Oregona. If NBA team need a player to earn three points, they should pick players from Villanova, and Oregona, and avoid the four colleges mentioned above. Players from UCLA, Iowa state, and Oregan tend to have higher assist percentage, while players from stanford, Gergetown, and Maryland tend to lower assist percentage (close to 0). If NBA team need a player to earn points through corporation, they should pick players from UCLA, Iowa state, and Oregan. In conclusion, players from Oregona tend to have better performance while players from Maryland perform bad in statistics.

```{r}
df_draft <-df_processed %>%
  group_by(draft_round)%>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))%>%
  select(draft_round, average_total_rebounds:win_shares)%>%
  pivot_longer(cols=average_total_rebounds:win_shares)
df_draft$draft_round<-as.character(df_draft$draft_round)
ggplot(df_draft, aes(x=name, y=value, fill=draft_round))+
  geom_bar(stat="identity", position = position_dodge())

```
```{r}
#install.packages("ggforce")
library(ggforce)
p<-ggplot(df_processed, aes(x = player_height, y = player_weight)) +
    geom_point(alpha=0.3, size=1) +
    labs(x = "Height", y = "Weight", title = "Scatter plot by Draft Year")
for (n in c(1,2, 3)){
  p <- p + facet_wrap_paginate(~draft_year, ncol = 4, nrow = 4, page = n)
  print(p)
}
```
